Size class small comparisons

### 2020-08-05 analysis/reports/smallness.Rmd
gleanings from shiny

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(drake)
library(dplyr)
library(ggplot2)

@@

###### Skew for small sites
```{r skewness small, fig.dim = c(4, 3)}

ggplot(filter(small_di, s0 >2), aes(x = skew_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Skew, <=")

ggplot(filter(small_di, s0 >2), aes(x = skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Skew, <")

small_di %>%
  filter(s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high_ltet = mean(skew_percentile > 95),
            prop_skew_high_lt = mean(skew_percentile_excl > 95),
            n_skew_sites = dplyr::n())

@@

###### Evenness
```{r evenness result small, fig.dim = c(4, 3)}

ggplot(filter(small_di), aes(x = simpson_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Even, <=")

ggplot(filter(small_di), aes(x = simpson_percentile_excl)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Even, <")

small_di %>%
  group_by(fia_yn) %>%
  summarize(prop_even_low_ltet = mean(simpson_percentile < 5),
            prop_even_low_lt = mean(simpson_percentile_excl < 5),
            n_even_sites = dplyr::n())

@@

For these comparisons, 
- For skewness, the effect is weak for both FIA and not-FIA (5% and 8% in the extremes, respectively)
- For evenness, the effect is much stronger for not-FIA (8 vs 15%)
- < vs <= matters quite a bit
    - Many of these sites can be quite small. Since we flagged a concern about having fewer than, arbitrarily, 100 elements, let's remove those:

###### Skew for small sites, removing very small
```{r skewness small no tiny, fig.dim = c(4, 3)}

small_di <- small_di %>%
  filter(nparts > 100)

ggplot(filter(small_di, s0 >2), aes(x = skew_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Skew, <=")

ggplot(filter(small_di, s0 >2), aes(x = skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Skew, <")

small_di %>%
  filter(s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high_ltet = mean(skew_percentile > 95),
            prop_skew_high_lt = mean(skew_percentile_excl > 95),
            n_skew_sites = dplyr::n())

@@

###### Evenness
```{r evenness result small no tiny, fig.dim = c(4, 3)}

ggplot(filter(small_di), aes(x = simpson_percentile)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Even, <=")

ggplot(filter(small_di), aes(x = simpson_percentile_excl)) +
  geom_histogram(bins = 100) +
  theme_bw() +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  ggtitle("Even, <")

small_di %>%
  group_by(fia_yn) %>%
  summarize(prop_even_low_ltet = mean(simpson_percentile < 5),
            prop_even_low_lt = mean(simpson_percentile_excl < 5),
            n_even_sites = dplyr::n())

@@

Removing sites with fewer than 100 elements in the feasible set

- *eliminates the difference between <= and <*
- We see a difference between FIA and other datasets 
    - For skew, 6 vs 12% in the extremes
    - For even, 10 vs 25% in the extremes

I am skeptical of these comparisons. It may be that FIA is behaving qualitatively differently, but it may also be that, because the other datasets are less concentrated at *very small* FS, the effect of FS size still drives the difference even when we filter like this. On the strength of this filtering I'm not confident arguing *either* that FIA is qualitatively different, independent of size, or that these subsets are behaving the same.

I think a stronger comparison would be to break into size classes based on the number of parts.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(drake)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning = FALSE, message = FALSE) 
```

@@

Having played around with the shiny app (size_comparisons), here are some static plots that I think are faithful to what I glean from tinkering with cutoffs.

##### Trivially small

For a range of about 1-300 elements in the FS, I wouldn't argue for any kind of detectable pattern in any of the results:

```{r trivially small}

minparts = 1
maxparts = 300

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts,
              s0 > 2), aes(skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts), aes(simpson_percentile)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

small_di %>%
  filter(nparts< maxparts, nparts > minparts,
         s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high = mean(skew_percentile_excl > 95),
            nskew_sites = dplyr::n())

small_di %>%
  filter(nparts< maxparts, nparts > minparts) %>% 
  group_by(fia_yn) %>%
  summarize(prop_even_low = mean(simpson_percentile < 5),
            neven_sites = dplyr::n())

@@

##### Small but not trivially small

There's one dynamic from a minimum of around 50 to 10000 elements, where:

- Not really anything detectable for skewness
- A weak effect for evenness for FIA, and quite a strong effect for evenness for other datasets

```{r medium small}

minparts = 50
maxparts = 10000

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts,
              s0 > 2), aes(skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts), aes(simpson_percentile)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

small_di %>%
  filter(nparts< maxparts, nparts > minparts,
         s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high = mean(skew_percentile_excl > 95),
            nskew_sites = dplyr::n())

small_di %>%
  filter(nparts< maxparts, nparts > minparts) %>% 
  group_by(fia_yn) %>%
  summarize(prop_even_low = mean(simpson_percentile < 5),
            neven_sites = dplyr::n())

@@

##### Big

And then anything over about 1500 elements:

- Effects for both skewness and evenness, which only get stronger as you increase the minimum
- Consistently more pronounced for other datasets for FIA, but the difference decreases as you increase the minimum

```{r big}

minparts = 1500
maxparts = exp(30)

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts,
              s0 > 2), aes(skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts), aes(simpson_percentile)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

small_di %>%
  filter(nparts< maxparts, nparts > minparts,
         s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high = mean(skew_percentile_excl > 95),
            nskew_sites = dplyr::n())

small_di %>%
  filter(nparts< maxparts, nparts > minparts) %>% 
  group_by(fia_yn) %>%
  summarize(prop_even_low = mean(simpson_percentile < 5),
            neven_sites = dplyr::n())

@@

##### Really big

Such that, over about 10000 elements, FIA looks similar to other datasets.

```{r really big}

minparts = 10000
maxparts = exp(30)

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts,
              s0 > 2), aes(skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

ggplot(filter(small_di,
              nparts< maxparts, nparts > minparts), aes(simpson_percentile)) +
  geom_histogram(bins = 100) +
  facet_wrap(vars(fia_yn), scales = "free_y") +
  theme_bw()

small_di %>%
  filter(nparts< maxparts, nparts > minparts,
         s0 > 2) %>%
  group_by(fia_yn) %>%
  summarize(prop_skew_high = mean(skew_percentile_excl > 95),
            nskew_sites = dplyr::n())

small_di %>%
  filter(nparts< maxparts, nparts > minparts) %>% 
  group_by(fia_yn) %>%
  summarize(prop_even_low = mean(simpson_percentile < 5),
            neven_sites = dplyr::n())

@@
